Discussion Draft

### 2020-04-02 doc/Discussion.rmd
small clean up

@@

[@laneKauppi2019]

The common direction of heterochiasmy is female biased, however male biased heterochiasmy has significant number of observations (list). Our results are novel in the sense that the evolutionary scale of the male-specific is short.

The prediction from two of the modifier models, haploid selection and SACE are that males will generally evolve lower recombination rates. <Male biased heterochiasmy predicted XX> The two locus modifier model  does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).

There are no prediction for male biased heterochiasmy evolution direction under the COM <pairing model>, but it should be noted that this model was not't designed to incorporate this type of data. This model were design to explain the positive interference pattern for of the recombination landscape by incorporating a cell mechanics of rapid prophase movements (RPM) that are thought to facilitate homolog pairing in many species.

Under the spindle based model faster male specific evolution of the genome wide recombination rate is primarily driven by differences in the strength of the spindle assemby checkpoint (SAC). (we won't go deeply into the molecular details (but the genes and pathways are will known, we refer the reader to [@laneKauppi2019]). The transition from metaphse to analphase is irreversable, so all bivalents in the cell has to be set up correctly before anaphase to prevent lower fitness later in the gametogenesis pathway. If unstable bivalents are detected on the MI spindle, the SAC delays the transition and buys time for the lose bivalent to correctly attach to the MT spindle. If this is delayed long enough, apoptosis will be triggered. If there are genetic variants which increase the number of unstable bivalents, they selected against in favor to genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

It is known that the strength of the SAC varies between the forms of gametogenesis.
(via the rate of anueploidy or achiasmate bivalents in mature gametes). 
Sperm have lower anueploidy rates compared to oocytes (other lines of evidence) indcating that

the SAC in sperm is more strict compared to oocytes (which have been discribed as 'leaky'). In sperm single unpaired bivalents can trigger the SAC. While in females more oocytes progress through meiosis even with unpaired bivalents.

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volumne (cite) -- which result in differing morphology of the metaphase MT spindle.

<leaky oocyte SAC linked to acentrosome spindle (mulitple aMTOCs) and cell volumne, these are features which differ in spermatocyttes (small, centrosome spindle, 2MTOCs)>

@@

```{r within.mouse.var, include=FALSE}
#2. Greater between cell variation in females, leaky SAC in oocytes can cause more between cell variation within an individual.
#2a. relaxed selection on chromosome structure in oocyte / leaky SAC, can lead to larger variance in crossover number between cells. The opposite dynamics of the purifying selection for 'strict' SAC in spermatocytes.

@@

**the within animal variation is equvillent to between cell variation**
[@lynn2002]

Our results of greater between cell variation . these results are supported by human and mouse (Gruhn (Kong)?  (Zwel -- Cole).  These data are also scant, between cell variance isn't often measured especially for both sexes. (given they require single cell observations for both sexes).

While the modifier models parametrize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant. (selection on penetrance).

@@

As mentioned above (fundamental/conserved) sex differences in the MI spindle presence of centrioles in males anchors spindle at two cell poles. While oocytes lack centrioles and so the MT spindle is more difuse and made of mulitple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007). This or larger cell volumne might -- lead to anaphase promoting signal being more difuse (or tension signal?)

2. Greater between cell variation in females, leaky SAC in oocytes can cause more between cell variation within an individual.

2a. relaxed selection on chromosome structure in oocyte / leaky SAC, can lead to larger variance in crossover number between cells. The opposite dynamics of the purifying selection for 'strict' SAC in spermatocytes.

2b. (leaky SAC could be adaptive/prevent to selfish genetic elements (Lampson))?

@@

Our results on the broad scale pattern of crossover placement are generally supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale landscape is a highly conserved trait.

The two locus modifier model and SACE model make predictions for sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.

@@

Just as the difference in sac (checkpoint) strength can lead to disinct evolutionary trajectories of the genome wide recombiation rate for oocytes and spermatocytes, the other side of the same coin is that there is increased variance between cells for crossover number. (weaker / leaky checkpoint )

@@

0.  Our results from MSM and PWD show that the sc length compaction is independant (of the direction of heterochiasmy. Two strains have have more crossovers in males, but females still have longer sc axis. _This indecates that the sexual dimorphism in chromosome organization / DNA compaction is a conserved part of recombination landscape / heterochiasmy_

The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002], [@gruhn2013].

Very few models make predictions for how chromatin organization relates to variation in recombination rates / the recombination landscape.  (the chromatin organization is how chromosomes with the same number of base pairs) are transformed into bottle brush structures with a central axis and most of the genetic material arranged as loops emanating outwards (figure). 
Only the COM model states that chromatin structure will be longer in oocytes due to greater cell volume.  This prediction / model might fit broader pattern (such as in At where pollen is the larger cell and has longer axis length) [@cahoonLibuda2019].  
Predictions for this trait are  inconclusive (due to lack of cytology measures from both sexes. We argue that the chromatin structure (and axis length) should be considered as an integral parameter when studying recombination landscape.  The sex differences in chromatin strucutre are closely connected to sex differences  in the interference strength and our results mirror those in Petkov et al.  (prediction, shorter-loop will have stronger interference).

According to the tethered loop model of DSB formation (cite) -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**.  Only one of the models COM aligns with our observed results for sexual dimorphism for axis length. **COM** predicts that axis length will be larger in oocytes due to the larger volume.

@@

Evolved traits, Evolution of interference (hardest section)
(interference is a fundamental aspect of the recombination landscape)
Some basics of the constraints bounds of interference are known, yet there are still many unknowns regarding the relationship of genome wide recombination rate and interference strength.  The non-random spacing of crossovers along the 2D length of chromosomes is the first level of quantification of crossover interference (cite gamma COI papers). Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength. This pattern has empirical support from the most species (Otto and Payseur) and fits well with  the fundamental relationship between the SC area or axis length, the physical upper limit for the number of crossovers.
Our results support an opposite pattern, a positive correlation between interference strength and genome wide recombination rates; we find support that interference strength has evolved in the two groups of male strains. There is a small number of positive correlations between genome wide recombination rate and interference strength in the literature. The within sex comparison of two breeds of cattle with different genome wide recombintion rates (Ma et al), between lab and wild mice of Peromyscus leucopus, and in a previous house mouse cross (Dumont,  preliminary data (HVR unpublished).
Theoretical models haven't really considered the evolution of interference strength, neither the haploid selection or two locus modifier model can not be applied to evolution of interference strength. While the SACE modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength. The COM model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, it lacks evolutionary based predictions. We propose the spindle based selection model would support the evolution of interference strength in the positive direction via modulation of the amount of sister cohesion connecting homologs (figure).
Models from Goldstein et al (review in Otto and Payseur) suggest that if this pattern is widespread interference evolves whenever increased recombination rates evolve. Perhaps a distinguishing feature of models which come to this finding is that the number of crossovers is kept constant. The space across multiple loci (veller?) or between multiple crossovers increases in a positive manner with genome wide recombination rates. Given that the empirical range of crossovers per chromosome is quite small (1-3  (Otto Payseur 2019, Stapley et al 207) and  the obligate crossover rule, the assumption of constraining the number of crossovers per chromosome fits well with empirical data.

>>>>

#our results of reversed hetC are novel due to the short evolutionary disances across the panel of mouse strains

@@

The common direction of heterochiasmy is female biased, however male biased heterochiasmy examples are not especially rare. Our results are novel from previous examples of male biased heterochiasmy species because the panel of strains have much shorter evolutionary distances. This context gives us insights to sex-specific evolution of recombination rates.

The prediction from two of the modifier models, haploid selection and SACE, state that males will generally evolve lower recombination rates. Male biased heterochiasmy is predicted under the two locus modifier model when XXXX.  <does parameterize cases for male biased heterochiasmy, dependent on the stage of action for the driver (MI or MII) and the linkage status (see table 1 Brandvain and Coop).>

There are no prediction for male biased heterochiasmy evolution direction under the COM <pairing model>, but it should be noted that this model was not designed to incorporate this type of data. This model was design to explain the positive interference pattern for of the recombination landscape by incorporating a cell mechanics of rapid prophase movements (RPM) that are thought to facilitate homolog pairing in many species.

Under the spindle based model faster male evolution in male genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) <acting on the chromosome structure>.

The genes and pathways are will known, we refer the reader to [@laneKauppi2019].

The transition from metaphase to anaphase is irreversible, so all bivalents in the cell have to be aligned correctly to prevent anueploidy.

Unstable bivalents are detected on the MI spindle, the SAC delays the cell cycle which potentially allows time for lose bivalents to correctly attach to the MT spindle. If the 'unattached' persists apoptosis will be triggered. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Thus the SAC can lead to stabilizing selection on the chromosome structure.

Sperm have lower anueploidy rates compared to oocytes (other lines of evidence) indicating that the 'strength' of the SAC varies between oogenesis and spermatogenesis. In sperm single unpaired bivalents can trigger the SAC. While in females more oocytes progress through meiosis even with unpaired bivalents (refereed to as leaky SAC).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite) which result in differing morphology of the metaphase microtubule spindle.

<leaky oocyte SAC linked to acentrosome spindle (multiple aMTOCs) and cell volume, these are features which differ in spermatocyttes (small, centrosome spindle, 2MTOCs)>

@@

```{r within.mouse.var, include=FALSE}
#review results, other data

#2. Greater between cell variation in females, leaky SAC in oocytes can cause more between cell variation within an individual.

#2a. relaxed selection on chromosome structure in oocyte / leaky SAC, can lead to larger variance in crossover number between cells. The opposite dynamics of the purifying selection for 'strict' SAC in spermatocytes.

@@

Our results of greater between cell variation in females, these results are supported by human and mouse (Gruhn, Lynn Kong). [@lynn2002].

These data are also scant, between cell variance isn't often measured especially for both sexes. (given they require single cell observations for both sexes).

While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant. (selection on penetrance).

@@

The opposite dynamics of the purifying selection for the 'strict'  SAC in spermatocytes is relaxed selection from a 'permissive' SAC. As mentioned above a fundamental sex difference in the metaphase spindle is the presence of centrioles in spermatocytes which anchors and nucleates the microtubulue spindle at the two cell poles. While oocytes lack centrioles which results in a difuse network of microtubules with mulitple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007). This or larger cell volumne may lead/be a potnetial source of the leaky SAC in oocytes. Leaky SAC, relaxed selection on the anaphase to metaphase transition could increase the between cell variance for crossover number

@@

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait.

The **two locus modifier** and **SACE** model make predictions for sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.

@@

Just as the difference in SAC (checkpoint) strength can lead to disinct evolutionary trajectories of the genome wide recombination rate for oocytes and spermatocytes, the other side of the same coin is that there is increased variance between cells for crossover number. (weaker / leaky checkpoint )

@@

Our results from MSM and PWD show that the sc length compaction is independant (of the direction of heterochiasmy. Two strains have have more crossovers in males, but females still have longer sc axis. **This indecates that the sexual dimorphism in chromosome organization / DNA compaction is a conserved part of recombination landscape / heterochiasmy**

The results of sexual dimorphism in chromatin structure is supported by the literature, especially for mammals [@lynn2002, @gruhn2013].

Very few models make predictions for how chromatin organization relates to variation in recombination rates / the recombination landscape.  (the chromatin organization is how chromosomes with the same number of base pairs) are transformed into bottle brush structures with a central axis and most of the genetic material arranged as loops emanating outwards (figure). 
Only the COM model predicts sexual dimorphism in chromatin structure will be longer in oocytes due to greater cell volume.  This prediction / model might fit broader pattern (such as in At where pollen is the larger cell and has longer axis length) [@cahoonLibuda2019].

Predictions for this trait are  inconclusive (due to lack of cytology measures from both sexes. We argue that the chromatin structure (and axis length) should be considered as an integral parameter when studying recombination landscape.

We also hypothesize that the sex differences in chromatin structure are closely connected to sex differences  in the interference strength and our results mirror those in [@petkov2007] and follow the prediction that shorter-loop will have stronger interference.

According to the tethered loop model of DSB formation (cite schmenti?), the 'real estate' for COs are the chromatin / DNA loops, the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**.  Only one of the models COM aligns with our observed results for sexual dimorphism for axis length. **COM** predicts that axis length will be larger in oocytes due to the larger volume.

@@

Interference is a fundamental aspect of the recombination landscape. Some basics of the constraints bounds of interference are known, yet there are still many unknowns regarding the relationship of genome wide recombination rate and interference strength.  The non-random spacing of crossovers along the 2D length of chromosomes is the first level of quantification of crossover interference (cite gamma COI papers). Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength. This pattern has empirical support from the most species (Otto and Payseur) and fits well with  the fundamental relationship between the SC area or axis length, the physical upper limit for the number of crossovers.

Our results support an opposite pattern, a positive correlation between interference strength and genome wide recombination rates; we find support that interference strength has evolved in the two groups of male strains. There is a small number of positive correlations between genome wide recombination rate and interference strength in the literature. The within sex comparison of two breeds of cattle with different genome wide recombintion rates (Ma et al) [@ma2015_cattle], between lab and wild mice of Peromyscus leucopus, [@peterson2019].

Theoretical models haven't really considered the evolution of interference strength, neither the haploid selection or two locus modifier model can not be applied to evolution of interference strength. While the SACE modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength. The COM model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, it lacks evolutionary based predictions. We propose the spindle based selection model would support the evolution of interference strength in the positive direction via modulation of the amount of sister cohesion connecting homologs (figure).

Models from Goldstein et al (review in Otto and Payseur) suggest that if this pattern is widespread interference evolves whenever increased recombination rates evolve. Perhaps a distinguishing feature of models which come to this finding is that the number of crossovers is kept constant. The space across multiple loci (veller?) or between multiple crossovers increases in a positive manner with genome wide recombination rates. Given that the empirical range of crossovers per chromosome is quite small (1-3  (Otto Payseur 2019, Stapley et al 207) and  the obligate crossover rule, the assumption of constraining the number of crossovers per chromosome fits well with empirical data.
